{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming\\[0.25em]

### 2025-07-12 paper.tex
More editing and revising

@@

\subsubsection*{Two-Phase Methodology}
The technical focus of this paper fits within the interior mode of design science research \cite{adam-2021-dsr}.
To \emph{design} Denicek, we identify six \emph{formative examples} -- simple programming
tasks that manifest one or more of the desired programming experiences~(\S\ref{app:examples}).
Using those examples, we co-design the Denicek substrate and a simple web-based end-user programming
environment Webnicek~(\S\ref{sec:walk}), which is built directly on top of the substrate. Although
Webnicek can be used to complete end-user programming tasks, it is optimized for developing the
underlying substrate rather than for usability.

To \emph{evaluate} the usability of Denicek for the development of end-user programming systems,
we use it to build Datnicek (\S\ref{sec:case}), an innovative interactive data science notebook
that brings together a range of recent research work~\cite{kandel-2011-wrangler,drossos-2020-wrex,petricek-2022-thegamma,adams-2025-grove}.
We reflect on the development experience~(\S\ref{sec:case-reflection}) and provide a heuristic
evaluation of Denicek characteristics including importance and generality (\S\ref{sec:eval}).

\subsubsection*{Substrate}
The Denicek substrate brings together two central design ideas. First, it represents programs as
document trees consisting of nodes that can represent data, formulas, evaluated results, as well as
static content. Second, Denicek does not store the document tree itself, but instead, maintains
a sequence of edit operations through which the tree was constructed and transformed.

@@

The key insight that we present in this paper is that many compelling programming experiences can
be implemented by leveraging this uniform document representation alongside with a suitable
composition of the three primitive operations.

@@

@@

\noindent
To enable others build on top of Denicek, we share our compact open-source
implementation at: \url{https://github/removed/for/review} %d3sprog/denicek

@@

\section{Background}
\label{sec:background}
Denicek aims to support a class of systems associated with end-user programming \cite{myers-2006-eup},
notational freedom and self-sustainability~\cite{jakubovic-2023-techdims}, liveness and interactivity
\cite{horowitz-2023-qualities,rein-2019-live}. We see programming more as interacting with a medium
or a substrate \cite{kay-1977-media,klokmose-2015-webstrates} than writing text \cite{rpg-2012-revolution}.
We focus on systems that allow gradual progression from a user to a developer, as envisioned
in Smalltalk~\cite{reenskaug-1981-byte} and use the term end-user programming loosely to refer to a
part of this spectrum, also including spreadsheet systems and notebooks for data science.
% We first review related systems and their technical design (\S\ref{sec:background-sys})
% before discussing programming experiences (\S\ref{sec:background-exp}).

@@

\subsection{Programming Systems and Substrates}
\label{sec:background-sys}

\subsubsection*{Programming Systems and Substrates}
A number of systems illustrate the qualities Denicek aims to support and have a related internal
structure. Subtext, BootstrapLab and Infra~\cite{edwards-2005-subtext,jakubovic-2022-ladder,hall-2017-infra}
use structured document-based program representation and provide some of the desired programming
experiences on top of this representation. Many of those design ideas can be traced back to
Boxer~\cite{disessa-1986-boxer}, which introduced the \emph{naive realism} principle (what the user
sees is all there is) that we also follow in Webnicek.

@@

Merging of edits is most frequently done in version control systems such as git. Source control systems
based on a more rigorous design such as Pijul \cite{meunier-2024-pijul} delay merging to a later point
by using a graph or a lattice \cite{schurmann-2022-merging}. In the context of
programming environments, Grove \cite{adams-2025-grove} uses a commutative patches, with a graph
structure similar to Pijul, as the basis for a collaborative structure editor.

@@

@@

We review a range of compelling programming experiences explored in recent research,
focusing on those supported by Denicek.

>>>>

%\newcommand{\diff}[1]{{#1}}
%\newcommand{\note}[1]{}

@@

\subsubsection*{Two-Phase Methodology}
The technical focus of this paper fits within the interior mode of design science research \cite{adam-2021-dsr}.
To \emph{design} Denicek (Fig.~\ref{fig:teaser} (A)), we identify six \emph{formative examples} -- simple programming
tasks that manifest one or more of the desired programming experiences~(\S\ref{app:examples}).
Using those examples, we co-design the Denicek substrate and a simple web-based end-user programming
environment Webnicek~(Fig.~\ref{fig:teaser} (B)), which is built directly on top of the substrate. Although
Webnicek can be used to complete end-user programming tasks, it is optimized for developing the
underlying substrate rather than for usability.

To \emph{evaluate} the usability of Denicek for the development of end-user programming systems,
we use it to build Datnicek (Fig.~\ref{fig:teaser}~(C)), an innovative interactive data science notebook
that brings together a range of recent research work~\cite{kandel-2011-wrangler,drossos-2020-wrex,petricek-2022-thegamma,adams-2025-grove}.
\note{Clarify that Datnicek was designed after Denicek was completed.}
\diff{We reflect on the degree to which Denicek simplifies the implementation of a programming
system that is based on current research advances and was conceived after the design of Denicek
has been fully finalized~(\S\ref{sec:case-reflection}).} We also provide a heuristic evaluation
of Denicek characteristics including importance and generality (\S\ref{sec:eval}).

\subsubsection*{Computational Substrates}
Denicek brings together two central design ideas. First, it represents programs as
document trees consisting of nodes that can represent data, formulas, evaluated results, as well as
static content. Second, Denicek does not store the document tree itself, but instead, maintains
a sequence of edit operations through which the tree was constructed and transformed.

@@

\subsubsection*{Key Takeaways}
\note{Add more general takeaway (below) and clarify specific technical takeaway (here).}
\diff{The key insights of this paper are twofold. A specific technical takeaway is that many
compelling programming experiences can be implemented on top a uniform document representation,
by using a suitable composition of three primitive operations that operate on sequences of
edit operations.}

@@

\note{Add a more general takeaway suggested by Reviewer 3 (Thank you!)}
\diff{A more basic takeaway, illustrated by the development of Denicek, is that it is worth looking
for novel computational substrates that better support compelling programming experiences developed
in user-centric programming research.}

@@

\noindent
To enable others build on top of Denicek, we share our compact open-source
implementation at: \url{https://github.com/d3sprog/denicek}

@@

\section{Background}
\label{sec:background}
\note{Clarify focus to make paper more self-contained}
\diff{The premise of Denicek is that suitable structures for constructing programs can make it easier
to support a range of compelling user experiences. A prime example is
Smalltalk whose object-oriented structures
enable malleable programming experience~\cite{girba-2017-moldable,tcher-2019-malleable}.}

Denicek aims to support a class of systems associated with end-user programming, liveness and
interactivity \cite{myers-2006-eup,horowitz-2023-qualities,rein-2019-live}, notational freedom and
self-sustainability~\cite{jakubovic-2023-techdims}. We see programming as interacting with a medium
or a substrate \cite{kay-1977-media,klokmose-2015-webstrates,rpg-2012-revolution} and
use the term end-user programming loosely to refer to a part of this spectrum, also including
spreadsheet systems and notebooks for data science.

\note{Highlight related systems (discussed later) for better framing}
\diff{We follow systems such as BootstrapLab~\cite{jakubovic-2022-ladder}, which aims to support
gradual progression from a user to a developer envisioned in Smalltalk~\cite{reenskaug-1981-byte}
in a primarily graphical environment and Subtext~\cite{edwards-2005-subtext},
which develops a user-friendly programming environment based on object copying.}

@@

\subsection{Programming Systems}
\label{sec:background-sys}

\subsubsection*{Programming Systems and Substrates}
A number of systems illustrate the qualities Denicek aims to support and have a related underlying
structure. Subtext, BootstrapLab and Infra~\cite{edwards-2005-subtext,jakubovic-2022-ladder,hall-2017-infra}
use structured document-based program representation and provide some of the desired programming
experiences on top of this representation. Many of those design ideas can be traced back to
Boxer~\cite{disessa-1986-boxer}, which introduced the \emph{naive realism} principle (what the user
sees is all there is) that we also follow in Webnicek.

@@

Merging of edits is most frequently done in version control systems. The
Pijul system \cite{meunier-2024-pijul} delays merging to a later point
by using a graph or a lattice \cite{schurmann-2022-merging}. In the context of
programming environments, Grove \cite{adams-2025-grove} uses a commutative patches with a similar graph
structure as the basis for a collaborative structure editor. \diff{Grove patches
(construct, delete, relocate) are sufficient for a structure editor, but lack some structures and
operations that our formative examples rely on (collections, copying). However, the core model
of Grove provides a possible alternative basis for Denicek.}
\note{Added why Grove cannot be used directly by Denicek, but is interesting direction.}

@@

\note{Added report of our experience with what makes EUP systems hard to build.}
\subsubsection*{Building Programming Experiences}
\diff{
The challenge addressed by Denicek is that end-user programming experiences are hard to~build.
Our experience developing such systems \cite{petricek-2019-histogram,petricek-2022-thegamma,
jakubovic-2022-ladder,edwards-2005-subtext,edwards-2004-example} suggests two difficulties.
The first is the need to move between the concrete and the abstract \cite{jakubovic-2023-techdims}.
A sequence edits strikes a balance between those opposing representations.
The second is the need to combine multiple experiences, which requires
a major engineering effort when using a conventional program representation.}

% Wrangler~\cite{kandel-2011-wrangler}

% * Reviewers also ask about the concrete problem addressed (R2) and what makes end-user programming
% tools hard to build (R1). The latter question (aptly phrased by R1) deserves a deeper investigation.
% We believe the difficulty emerges when combining multiple programming experiences in a single system
% and from the need to move between the concrete and the abstract. This belief is based on our personal
% experience developing multiple systems (to be cited in the revision) and on a technical dimensions
% analysis (Sec. 8.2). A substrate based on edit history provides a middle ground between the
% abstract and the concrete and we show that it can be carefully designed to serve as the basis for a range of experiences.

@@

%We review a range of compelling programming experiences explored in recent research,
%focusing on those supported by Denicek.
